---
title: "An Exploration of NYC Home Mortgage Disclosure Act Data"
output: 
  html_document: 
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(maptools)
library(ggmap)
library(broom)
library(ggrepel)
library(statar)
library(sf)
library(treemap)
```

```{r, error = FALSE, message = FALSE, warning = FALSE, include=FALSE, cache = TRUE}
#loading data and making preliminary changes 
options("scipen"=100, "digits"=4)

hmda13_ny <- read_csv("hmda13_ny.csv")
hmda14_ny <- read_csv("hmda14_ny.csv")
hmda15_ny <- read_csv("hmda15_ny.csv")
hmda16_ny <- read_csv("hmda16_ny.csv")
hmda17_ny <- read_csv("hmda17_ny.csv")

hmda_ny_5yrs <- rbind(hmda13_ny, hmda14_ny, hmda15_ny, hmda16_ny, hmda17_ny) %>%
  mutate(census_tract_number = gsub("\\.", "", census_tract_number)) %>% filter(county_name %in% c("Bronx County", "Kings County", "Queens County", "New York County", "Richmond County"))

hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(outcome = ifelse(action_taken_name %in% 
                            c("Application approved but not accepted",
                              "Loan originated", "Loan purchased by the institution"),"Approved", 
                          ifelse(action_taken_name %in%
                                   c("Application denied by financial institution",
                                     "Preapproval request denied by financial institution"),
                                 "Denied", "Other")))

hmda_ny_5yrs$applicant_race_name_1[which(hmda_ny_5yrs$applicant_race_name_1 %in%
                                           c("Information not provided by applicant in mail, Internet, or telephone application",
                                             "Not applicable"))] <- "Unknown"


hmda_ny_5yrs <- hmda_ny_5yrs %>% mutate(race_alternative = applicant_race_name_1)
hmda_ny_5yrs$race_alternative[which(hmda_ny_5yrs$race_alternative %in% 
                                      c("American Indian or Alaska Native", "Native Hawaiian or Other Pacific Islander"))] <- "Other"

hmda_ny_5yrs$race_alternative[which(hmda_ny_5yrs$race_alternative == "Black or African American")] <- "Black"

hmda_ny_5yrs <-  hmda_ny_5yrs %>% mutate(income_bins = ifelse(applicant_income_000s %in% c(1:73), "Income Quartile 1", 
    ifelse(applicant_income_000s %in% c(74:108), "Income Quartile 2", 
        ifelse(applicant_income_000s %in% c(109:180), "Income Quartile 3", 
            ifelse(applicant_income_000s %in% c(181:133549), "Income Quartile 4",
                   "Unknown")))))

hmda_ny_5yrs <-  hmda_ny_5yrs %>% mutate(boro = 
        ifelse(county_name == "New York County", "Manhattan", 
            ifelse(county_name == "Kings County", "Brooklyn", 
              ifelse(county_name == "Queens County", "Queens", 
                     ifelse(county_name == "Bronx County", "Bronx",
                            ifelse(county_name == "Richmond County", 
                                   "Staten Island", NA))))))

census_tracts <- st_read("/Volumes/SPACESHIP/data_viz/data_scripts/visualizing-nyc-housing/2010_Census_Tracts", "geo_export_6e8e8432-67b3-4029-a002-fa59bbd3d5d3")

nta_shape <- st_read("/Volumes/SPACESHIP/data_viz/data_scripts/visualizing-nyc-housing/Neighborhood_Tabulation_Areas", "geo_export_f14927c2-d9cb-48d6-b521-916e745df67d")

rm(hmda13_ny, hmda14_ny, hmda15_ny, hmda16_ny, hmda17_ny)

crosswalk<- read_csv("crosswalk_nta_ct.csv")
crosswalk$X4 <- NULL

hmda_ny_5yrs <- hmda_ny_5yrs %>% left_join(crosswalk, by = c("census_tract_number" = "census_tract")) 

rm(crosswalk)
```


#Overview 

Homeownership is one of the most important ways of acculating wealth. In New York City, proprty values are high and thus homeownership is out of reach for many.Using Home Mortgage Disclosure Act data, I explore the demographics of mortgage applicants and the various outcomes of their applications in order to better understand who receives loans and who does not. Looking at how outcomes change over time and by group, allows us to better understand the way opportunities within the city are changing.

##Who is applying for mortgages? 

```{r, fig.width=10, fig.height=9, fig.align="center"}
#Loan applications are split pretty evenly between Manhattan, Brooklyn, and Queens, with White applicants representing a large share of the applications. The Unknown category is made up of applicants who did not disclose their race in their loan applications. Disparities in incomes between boroughs are shown in the second graph. 

#treemap for counties and race 
hmda_ny_5yrs %>% group_by(boro, race_alternative) %>% 
  summarise(n = n()) %>% 
  treemap(index = c("boro", "race_alternative"),
          vSize ="n", title = "Loan Applicants Organized by Borough and Race \nHMDA Data 2013-2017",
          fontsize.labels = c(20,15), 
          align.labels = list(c("left","top"),c("right","bottom")),
          palette = c( '#FD7400', '#BF4182', '#580F45', '#0D177F', '#1f8a70'),
          fontcolor.labels= '#F2F5F2', border.col = '#F2F5F2', 
          fontsize.title = 24)

#treemap of counties and income bins
hmda_ny_5yrs %>% group_by(boro, income_bins) %>% 
  summarise(n = n()) %>% 
  treemap(index = c("boro", "income_bins"),
          vSize ="n", fontsize.labels = c(25,15), 
          title = "Loan Applicants Organized by Borough and Income Quartiles \nHMDA Data 2013-2017",
          align.labels = list(c("left","top"),c("right","bottom")), 
          palette = c( '#FD7400', '#BF4182', '#580F45', '#0D177F', '#1f8a70'),
          fontcolor.labels= '#F2F5F2', border.col = '#F2F5F2', 
          fontsize.title = 24)

```

```{r, error = FALSE, message = FALSE, warning = FALSE}
#looking at differences in approval rates by male/female
hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(win_income = winsorize(applicant_income_000s, probs = c(.01, .99)))

ggplot(filter(hmda_ny_5yrs, applicant_sex_name %in% c("Female","Male")),
       aes(y=win_income, x=applicant_sex_name, fill=outcome), alpha = .3) +
  geom_boxplot() + scale_y_log10(expand=c(0,0)) + 
  scale_fill_manual(values = c('#1f8a70', '#E8291A', '#FFE11A')) + 
  theme(text = element_text(family = "Meiryo"), 
         panel.background = element_rect(fill = '#F2F5F2'),
        panel.grid.major = element_line(color = "#404040"), 
        panel.grid.minor = element_blank()) +
  labs(fill = "Outcome", x = "Gender of Applicant", 
       y = "Income (in $1,000s, Log Scale)", 
       title = "On Average, Male Applicants have Higher Incomes \nthan Female Applicants", 
       subtitle = "Income by Gender, grouped by Outcome", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017") 

#trying to add summary statistics 

violin_data <- filter(hmda_ny_5yrs, applicant_sex_name %in% c("Female","Male"))

ggplot(data = violin_data) +
  geom_violin(aes(y=win_income, x=applicant_sex_name, fill=outcome)) + 
  scale_y_log10(expand=c(0,0)) + 
  scale_fill_manual(values = c('#1f8a70', '#E8291A', '#FFE11A')) + 
  theme(text = element_text(family = "Meiryo"), 
         panel.background = element_rect(fill = '#F2F5F2'),
        panel.grid.major = element_line(color = "#404040"), 
        panel.grid.minor = element_blank()) +
  labs(fill = "Outcome", x = "Gender of Applicant", 
       y = "Income (in $1,000s, Log Scale)", 
       title = "On Average, Male Applicants have Higher Incomes \nthan Female Applicants", 
       subtitle = "Income by Gender, grouped by Outcome", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017") 

```