---
title: "hmda_explore.R"
output: 
  html_document: 
    code_folding: hide
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(maptools)
library(ggmap)
library(broom)
library(ggrepel)
library(statar)
library(sf)
```


```{r, error = FALSE, message = FALSE, warning = FALSE}
#loading data and making preliminary changes 
options("scipen"=100, "digits"=4)

hmda13_ny <- read_csv("hmda13_ny.csv")
hmda14_ny <- read_csv("hmda14_ny.csv")
hmda15_ny <- read_csv("hmda15_ny.csv")
hmda16_ny <- read_csv("hmda16_ny.csv")
hmda17_ny <- read_csv("hmda17_ny.csv")

hmda_ny_5yrs <- rbind(hmda13_ny, hmda14_ny, hmda15_ny, hmda16_ny, hmda17_ny) %>%
  mutate(census_tract_number = gsub("\\.", "", census_tract_number)) %>% filter(county_name %in% c("Bronx County", "Kings County", "Queens County", "New York County", "Richmond County"))

hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(outcome = ifelse(action_taken_name %in% 
                            c("Application approved but not accepted",
                              "Loan originated", "Loan purchased by the institution"),"Approved", 
                          ifelse(action_taken_name %in%
                                   c("Application denied by financial institution",
                                     "Preapproval request denied by financial institution"),
                                 "Denied", "Other")))

hmda_ny_5yrs$applicant_race_name_1[which(hmda_ny_5yrs$applicant_race_name_1 %in%
                                           c("Information not provided by applicant in mail, Internet, or telephone application",
                                             "Not applicable"))] <- "Unknown"


hmda_ny_5yrs <- hmda_ny_5yrs %>% mutate(race_alternative = applicant_race_name_1)
hmda_ny_5yrs$race_alternative[which(hmda_ny_5yrs$race_alternative %in% 
                                      c("American Indian or Alaska Native", "Native Hawaiian or Other Pacific Islander"))] <- "Other"

census_tracts <- st_read("/Volumes/SPACESHIP/data_viz/data_scripts/visualizing-nyc-housing/2010_Census_Tracts", "geo_export_6e8e8432-67b3-4029-a002-fa59bbd3d5d3")

```

##Home Mortgage Disclosure Act Data - Exploration in New York City 

Home Mortgage Disclosure Act data provides insight into which loans are approved and denied and includes details on loan characteristics such as use, amount, outcome, and on applicant characteristics such as race, income, and census tract. These graphs attempt to understand why some people receive loans while others do not, where most loans are approved and where most are denied, and more. 

```{r, error = FALSE, message = FALSE, warning = FALSE}
#number of loans apps by census tract
apps_by_tract <- hmda_ny_5yrs %>% 
  group_by(census_tract_number) %>% 
  summarise(total_apps = n())

#approvals by tract
approval_counts_by_tract <- hmda_ny_5yrs %>% 
  filter(outcome == "Approved") %>% 
  group_by(census_tract_number) %>% 
  summarise(total_approved = n())

approval_rates_by_tract <- 
  approval_counts_by_tract %>% 
  left_join(apps_by_tract, by = "census_tract_number") %>% 
  mutate(approval_rate = total_approved/total_apps)

census_tract_merged <- left_join(census_tracts, approval_rates_by_tract,
    by = c("ct2010" ="census_tract_number")) %>% 
      st_as_sf()

#dealing with ggplot
census_tract_merged$breaks <- cut(census_tract_merged$approval_rate, c(0, .2, .4, .6, .8, 1)) 

#plot- approval rates by census block #### 
ggplot() + 
  geom_sf(data = census_tract_merged,
               aes(fill = breaks), color = "white", lwd = 0) +
  labs(title = "Most census tracts have approval rates between 60% and 80%", 
       subtitle = "The highest approval rates are mostly found in Queens, while rates dip below 40% in historically lower-income areas such as East Brooklyn", 
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017") + 
  theme(axis.text=element_blank(),                      
        axis.title=element_blank(),                      
        panel.background = element_rect(fill = "#ffffff", color = NA),
        axis.line = element_blank(),
        legend.position = c(0.25,.75),
        text = element_text(family = "Meiryo"), 
        axis.ticks = element_blank()) + guides(fill=guide_legend(title="Approval Rate")) + scale_fill_manual(values = c('#ffffd4', '#fdd2c0', '#f3a683', '#e27a48', '#cc4c02'))

```


The map above and the next map create an understanding of loan approval rates across census tracts. While there are clearly differences in loan approval rates across the city, there is not a clear distinction between certain areas with higher approval rates, for example, there are high approval tracts in Brooklyn and the Bronx that are not know for being high income (as one might expect). In terms of low approval rates, this graph shows us that parts of East Brookyn, like Brownsville have lower rates than average. 

```{r, error = FALSE, message = FALSE, warning = FALSE, include=FALSE, eval = FALSE}
#approval rates by race by census tract
approval_counts_by_race <- hmda_ny_5yrs %>% 
  filter(outcome == "Approved") %>% 
  group_by(census_tract_number, applicant_race_name_1) %>% 
  summarise(total_approved = n())

df_for_chloropleth_race_approval <- hmda_ny_5yrs %>% 
  group_by(census_tract_number, applicant_race_name_1) %>% 
  summarise(total_apps = n()) %>% 
  left_join(approval_counts_by_race, 
            by = c("census_tract_number", "applicant_race_name_1")) %>%
  mutate(approval_rate = total_approved/total_apps)

df_for_chloropleth_black_approval <- hmda_ny_5yrs %>% 
  group_by(census_tract_number, applicant_race_name_1) %>% 
  summarise(total_apps = n()) %>% 
  left_join(approval_counts_by_race, 
            by = c("census_tract_number", "applicant_race_name_1")) %>%
  mutate(approval_rate = total_approved/total_apps) %>%
  filter(applicant_race_name_1== "Black or African American")

df_for_chloropleth_black_approval <- df_for_chloropleth_black_approval %>% 
  mutate(approval_rate_round = round(approval_rate, 2))

#upload shapefiles
census_tracts <- rgdal::readOGR("/Volumes/SPACESHIP/data_viz/data_scripts/visualizing-nyc-housing/2010_Census_Tracts", "geo_export_6e8e8432-67b3-4029-a002-fa59bbd3d5d3")

census_tract_merged_black <- sp::merge(x = census_tracts,
                                       y = df_for_chloropleth_black_approval,
                                       by.x ="ct2010", by.y= "census_tract_number", all.x = TRUE)

#dealing with ggplot
#tidy the spatial dataframe, make it into a dataframe
census_tracts_merged_black_df <- tidy(census_tract_merged_black)
#need to re-ad datat to it, by create a category it can merge with our data
census_tract_merged_black$polyID <- sapply(slot(census_tract_merged_black, "polygons"),
                               function(x) slot(x, "ID"))
#merge back with our data
census_tracts_merged_black_df <- merge(census_tracts_merged_black_df, 
                          census_tract_merged_black, 
                          by.x = "id", by.y="polyID")

census_tracts_merged_black_df$breaks <- cut(census_tracts_merged_black_df$approval_rate, c(0, .2, .4, .6, .8, 1))

#plot- black loan approval rates map ####
ggplot() + 
  geom_polygon(data = census_tracts_merged_black_df,
               aes(x = long, y = lat, group = group, fill = breaks)) + 
  scale_fill_brewer("Approval Rate", palette = "OrRd") +
  labs(title = "Highest loan approval rates for Blacks are in \nStaten Island and Queens", 
       subtitle = "When compared to approval rates that do not consider race, \nmost census tracts have lower approval rates for Blacks", 
       caption = "Home Mortgage Disclosure Act Data, CFPB") + 
  theme(line = element_blank(),                          
        axis.text=element_blank(),                      
        axis.title=element_blank(),                      
        panel.background = element_blank(), 
        text = element_text(family = "Meiryo")) + guides(fill=guide_legend(title="Approval Rate")) + scale_fill_manual(values = c('#ffffd4', '#fdd2c0', '#f3a683', '#e27a48', '#cc4c02'))

```

```{r}
#map of loan approval rates for low income applicants

census_tracts_bk <- census_tracts %>% filter(boro_name == "Brooklyn")

hmda_ny_5yrs <-  hmda_ny_5yrs %>% mutate(income_bins = ifelse(applicant_income_000s %in% c(1:73), "Income Quartile 1", 
                                             ifelse(applicant_income_000s %in% c(74:108), "Income Quartile 2", 
                                                    ifelse(applicant_income_000s %in% c(109:180), "Income Quartile 3", 
                                                           ifelse(applicant_income_000s %in% c(181:133549), "Income Quartile 4", NA)))))

apps_outcome_Q1_bk <- hmda_ny_5yrs %>% filter(income_bins == "Income Quartile 1") %>% group_by(census_tract_number, as_of_year, outcome) %>% summarise(outcome_tot = n())

apps_total_Q1_bk <- hmda_ny_5yrs %>% filter(income_bins == "Income Quartile 1") %>% group_by(census_tract_number, as_of_year) %>% summarise(tot_apps = n())

approval_rate_5yr_changes <- apps_outcome_Q1_bk %>% left_join(apps_total_Q1_bk, by = c("census_tract_number", "as_of_year")) %>% filter(outcome == "Approved", as_of_year %in% c(2013, 2017)) %>% mutate(approval_rate = outcome_tot/tot_apps)

approval_rate_5yr_changes$prior <- lag(approval_rate_5yr_changes$approval_rate)

census_tract_merged <- approval_rate_5yr_changes %>% filter(as_of_year == 2017) %>% mutate(approval_rate_change = approval_rate- prior) %>% right_join(census_tracts_bk,  by = c("census_tract_number" = "ct2010"))

census_tract_merged$breaks <- cut(census_tract_merged$approval_rate_change, c(-0.8, -0.3, 0, 0.3, 0.8))


ggplot() + geom_sf(data = census_tract_merged, 
                   aes(fill = breaks), color = "white", lwd = 0) + guides(fill=guide_legend(title="Change in Approval Rate \nbetween 2013-2017")) +  scale_fill_manual(values = c('#cc4c02', '#e27a48', "#a6d96a", "#1a9641")) + theme(axis.text=element_blank(),                      
        axis.title=element_blank(),                      
        panel.background = element_rect(fill = "#ffffff", color = NA),
        axis.line = element_blank(),
        text = element_text(family = "Meiryo"), 
        axis.ticks = element_blank()) + labs(title = "Gentrifying neighborhoods such as Bed-Stuy, Greenpoint, and Crown Heights show \nsignificant decreases in loan approval rates for low-income applicants over the last five years", subtitle = "However, many areas further from Manhattan show increased approval rates")

```

```{r}
#map that looks at change in number of loan applications
#map that groups by NTA rather than census tract - merge early 

hmda_ny_5yrs %>% filter(income_bins == "Income Quartile 1") %>% group_by(census_tract_number, as_of_year) %>% summarise(tot_apps = n()) %>% filter(as_of_year %in% c(2013, 2017)) 

hmda_ny_5yrs %>% group_by(census_tract_number, as_of_year) %>% summarise(tot_apps = n()) %>% filter(as_of_year %in% c(2013, 2017))

#merge then group by NTA 



```

This map looks at approval rates for Black and American American loan applicants by census tracts. When compared to the previous graph, it is clear the Black approval rates are much lower than the average approval rates in most of the census tracts. Many of the tracts that had rates between 50-75 percent, are now in the 25-50 percent range. This indicates that Black applicants receive below average loan approval rates in many neighborhoods. Understanding the cause of this outcome would be helpful in improving socioceonomic equality, as homeownership is an important step towards gaining wealth. 

```{r, error = FALSE, message = FALSE, warning = FALSE}
apps_per_year_outcome_race <- hmda_ny_5yrs %>% 
  group_by(as_of_year, race_alternative, outcome) %>%
  summarise(apps_per_year_outcome_race = n())

apps_per_year_outcome <- hmda_ny_5yrs %>% 
  group_by(as_of_year, race_alternative) %>% 
  summarise(apps_per_year_outcome = n())

#approval rates over the five years ####
approval_rate_year_race <- apps_per_year_outcome_race %>% 
  left_join(apps_per_year_outcome, 
            by = c("as_of_year", "race_alternative")) %>% 
  mutate(outcome_rate = apps_per_year_outcome_race/apps_per_year_outcome) %>%
  filter(outcome == "Approved") 

ggplot(approval_rate_year_race, 
       aes(x = as_of_year, y = outcome_rate, color = race_alternative)) +
  geom_point() + geom_line() + 
  scale_x_continuous(breaks = c(2013, 2014, 2015, 2016, 2017)) + 
  labs(title = "Loan approval rates for Blacks are consistently below Whites and Asians", x = "Year", y = "Approval Rate", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017") + 
  geom_text_repel(data = filter(approval_rate_year_race, as_of_year == 2017),
                  aes(x = 2017, label=race_alternative)) + 
  theme(legend.position = "none", text = element_text(family = "Meiryo"), panel.background = element_rect(fill = '#F2F5F2'), panel.grid.minor = element_line(linetype = "dashed"), axis.ticks = element_blank()) + scale_color_manual(values = c('#E8291A','#FD7400', '#004358', '#1f8a70', '#BEDB39'))
```


I was also curious whether loan approval rates by race have changed over the last five years. This graph shows that Asians and Whites have consistently had the highest approval rates. The unknown category is made up of people who did not choose the dislose their race in their mortgage application, however, given that they have the highest approval rates, there is a large chance that this group is made up mostly of White applicants. Most races have had consistent approval rates over the last five years, however, American Indians and Native Alaskans saw a sharp decrease in their loan approval rates in 2016, however their rates have returned to the pre-2016 average. This graph displays important differences in racial approval rates and inspires further research on whether these differences are due to historical differences in wealth or racial discrimination in lending. 

```{r, error = FALSE, message = FALSE, warning = FALSE}
#denial reasons over time
denial_reasons_year <- hmda_ny_5yrs %>% filter(outcome == "Denied", !is.na(denial_reason_name_1)) %>% 
  group_by(as_of_year, denial_reason_name_1) %>% summarise(count_per_reason = n())

denial_count_year <- hmda_ny_5yrs %>% filter(outcome == "Denied", !is.na(denial_reason_name_1)) %>% 
  group_by(as_of_year) %>% summarise(tot_denied = n())

#denial reasons changing over time ####
denial_reasons_year %>% 
  left_join(denial_count_year, by = "as_of_year") %>% 
  mutate(denial_rate = count_per_reason/tot_denied) %>% 
  ggplot(aes(x = as_of_year, y = denial_rate, color = denial_reason_name_1)) + 
  geom_point() + geom_line() + labs(title = "Debt-to-Income Ratio is consistently the most \nfrequent reason for denial", 
                                    subtitle = "Insufficient Collateral and Incomplete Applications have been decreasing as \ndenial reasons while Credit History-related denials have been increasing", 
                                    caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017", 
                                    x = "Year", y = "Denial Rate", color = "Denial Reason") + scale_color_manual(values = c('#E8291A','#FFE11A', '#004358', '#1f8a70', '#BEDB39',  '#FD7400', '#BF4182', '#580F45', '#0D177F')) + theme(text = element_text(family = "Meiryo"), panel.background = element_rect(fill = '#F2F5F2'), axis.ticks = element_blank())
```


In order to better understand why different races are denied loans, I look at denial reasons over time, and differences in denial reasons for different races. The following graph explores whether and how denial reasons have changed over time. This graph shows that an insuffiencient debt-to-income ratio is the highest frequency reason for denial. A determination that the property of the value is not high enough to serve as collateral for the requested loan amount ties for the second most frequent reason for denial, along with a poor credit history. While the differences in these rates of denial are minor, the highest frequcny rate suggests that many people are applying for mortgages that the banks believe they are unable to afford on their current incomes. Is this because the property values are high and people want to remain in a certain area or is this because people are simply attracted to things outside of their price ranges? These are some of the research questions raised by the graph. 


```{r, error = FALSE, message = FALSE, warning = FALSE, fig.height=6}
num_denied_by_race <- hmda_ny_5yrs %>% 
  filter(outcome == "Denied", !is.na(denial_reason_name_1)) %>%
  group_by(applicant_race_name_1) %>% 
  summarise(tot_apps_denied = n())

denial_reasons_race <- hmda_ny_5yrs %>% 
  filter(outcome == "Denied", !is.na(denial_reason_name_1)) %>% 
  group_by(applicant_race_name_1, denial_reason_name_1) %>% 
  summarise(denied_by_reason = n())

#denial reasons rates race ####
denial_reasons_race %>% 
  left_join(num_denied_by_race, by = "applicant_race_name_1") %>%
  mutate(denial_rate = denied_by_reason/tot_apps_denied) %>% 
  ggplot() + 
  geom_bar(aes(x = applicant_race_name_1, 
               y = denial_rate, 
               fill = denial_reason_name_1),
           stat = "identity") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), text = element_text(family = "Meiryo"), panel.background = element_rect(fill = '#F2F5F2'), axis.ticks = element_blank()) + 
  labs(title = "Credit History and Debt-to-Income Ratio are the two \nmost frequent loan denial reasons", 
       subtitle = "However, the distribution of reasons differ significantly by race", 
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017", 
       x = "Applicant Race", y = "Denial Rate", fill = "Denial Reason") + scale_fill_discrete(name = "Denial Reason")  + scale_fill_manual(values = c('#E8291A','#FFE11A', '#004358', '#1f8a70', '#BEDB39',  '#FD7400', '#BF4182', '#580F45', '#0D177F'))
```


Having seen the overall denial rates, the following graph attempts to visualize how the racial breakdown of denial reasons differs from the averages. Asians and Whites are very rarely denied because of their credit histories while Blacks, Native Americans, and Pacific Islanders are most frequently denied because of their credit histories. This begs a better understanding of why certain racial groups would have worse credit histories, whether this measure if being used to deny people because it is more subjective than other measures, or what can be done to improve the credit histories of historically disadvantaged groups.

```{r, eval = FALSE, include = FALSE}
library(packcircles)

counts_race <- hmda_ny_5yrs %>% 
  group_by(applicant_race_name_1) %>% 
  summarise(tot_apps_race = n())

#approved
approved_race_rate <- hmda_ny_5yrs %>% 
  group_by(outcome, applicant_race_name_1) %>%
  summarise(n = n()) %>%
  left_join(counts_race, by = "applicant_race_name_1") %>% 
  mutate(rate= n/tot_apps_race) %>% filter(outcome == "Approved")

approved_race_rate <- as.data.frame(approved_race_rate)
packing <- circleProgressiveLayout(approved_race_rate$rate, sizetype = 'area')
approved_race_rate <- cbind(approved_race_rate, packing)
dat.gg <- circleLayoutVertices(packing, npoints =50 )
dat.gg$id[which(dat.gg$id == 1)] <- "American Indian or Alaska Native"
dat.gg$id[which(dat.gg$id == 2)] <- "Asian"
dat.gg$id[which(dat.gg$id == 3)] <- "Black or African American"
dat.gg$id[which(dat.gg$id == 4)] <- "Native Hawaiian or Other Pacific Islander"
dat.gg$id[which(dat.gg$id == 5)] <- "Unknown"
dat.gg$id[which(dat.gg$id == 6)] <- "White"

ggplot() + 
  geom_polygon(data = dat.gg, aes(x,y, group = id, fill = as.factor(id)), color = "black") +
  geom_text_repel(data = approved_race_rate, aes(x,y,label = applicant_race_name_1)) + 
  scale_size_continuous(range = c(1,4)) + labs(title = "Denied") + 
  labs(title = "White and Asian applicants receive higher rates of loan approval than other races",
       subtitle = "Areas represent the percent of loan application approved within racial groups", 
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017", fill = "Race") + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        axis.title.x = element_blank(), axis.title.y = element_blank(), 
        axis.ticks = element_blank(), axis.text = element_blank(), text = element_text(family = "Meiryo")) + 
  scale_fill_manual(values = c('#E8291A','#FFE11A', '#004358', '#1f8a70', '#BEDB39',  '#FD7400'))


#denied
denied_race_rate <- hmda_ny_5yrs %>% 
  group_by(outcome, applicant_race_name_1) %>%
  summarise(n = n()) %>%
  left_join(counts_race, by = "applicant_race_name_1") %>% 
  mutate(rate= n/tot_apps_race) %>% filter(outcome == "Denied")

denied_race_rate <- as.data.frame(denied_race_rate)
packing <- circleProgressiveLayout(denied_race_rate$rate, sizetype = 'area')
denied_race_rate <- cbind(denied_race_rate, packing)
dat.gg <- circleLayoutVertices(packing, npoints =50 )
dat.gg$id[which(dat.gg$id == 1)] <- "American Indian or Alaska Native"
dat.gg$id[which(dat.gg$id == 2)] <- "Asian"
dat.gg$id[which(dat.gg$id == 3)] <- "Black or African American"
dat.gg$id[which(dat.gg$id == 4)] <- "Native Hawaiian or Other Pacific Islander"
dat.gg$id[which(dat.gg$id == 5)] <- "Unknown"
dat.gg$id[which(dat.gg$id == 6)] <- "White"

ggplot() + 
  geom_polygon(data = dat.gg, aes(x,y, group = id, fill = as.factor(id)), color = "black") +
  geom_text_repel(data = denied_race_rate, aes(x,y,label = applicant_race_name_1)) + 
  scale_size_continuous(range = c(1,4)) +
  labs(title = "Historically Disadvantaged Racial Groups Have Higher Rates of Denial",
       subtitle = "Areas represent the percent of loan application denied within racial groups", 
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017", fill = "Race") + 
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), panel.grid.minor = element_blank(), 
        axis.title.x = element_blank(), axis.title.y = element_blank(), 
        axis.ticks = element_blank(), axis.text = element_blank(), text = element_text(family = "Meiryo"), panel.background = element_rect(fill = '#F2F5F2')) + 
  scale_fill_manual(values = c('#E8291A','#FFE11A', '#004358', '#1f8a70', '#BEDB39', '#FD7400'))

```



```{r, eval = FALSE, include = FALSE}
hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(win_loan_amount = winsorize(loan_amount_000s, probs = c(.01, .99)))

ggplot(hmda_ny_5yrs, aes(x = win_loan_amount, fill = county_name)) +
  geom_histogram(alpha = .5, bins = 100) + 
  scale_x_continuous(breaks= c(0, 500, 1000, 1500, 2000, 2500, 3000, 3500, 4000, 4500)) +
  labs(title = "Most loans applications are made for under $1 million, \n and most of those made over $1 million are in New York County or Kings County", 
       subtitle = "The distribution of loan application amounts by county shows large spikes around $400,000", 
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017", 
       fill = "County", x = "Loan amount (in $1,000s)", y = "Number of Loan Applications") + scale_fill_manual( values = c('#E8291A','#FFE11A', '#004358', '#1f8a70', '#BEDB39',  '#FD7400', '#BF4182', '#580F45')) + theme(legend.position = c(.9, .7)) + theme(text = element_text(family = "Meiryo"), panel.background = element_rect(fill = '#F2F5F2'))
```

```{r, error = FALSE, message = FALSE, warning = FALSE}
#looking at differences in approval rates by male/female
hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(win_income = winsorize(applicant_income_000s, probs = c(.01, .99)))

ggplot(filter(hmda_ny_5yrs, applicant_sex_name %in% c("Female","Male")),
       aes(y=win_income, x=applicant_sex_name, fill=outcome), alpha = .3) +
  geom_boxplot() + scale_y_log10(expand=c(0,0)) + 
  scale_fill_manual(values = c('#1f8a70', '#E8291A', '#FFE11A')) + 
  theme(text = element_text(family = "Meiryo"), 
         panel.background = element_rect(fill = '#F2F5F2'), panel.grid.major = element_line(color = "#404040"), panel.grid.minor = element_blank()) +
  labs(fill = "Outcome", x = "Gender of Applicant", y = "Income (in $1,000s)", 
       title = "There is a difference in the median income of \nmen and women who are approved for loans", 
       subtitle = "While approval relies on other factors, the distribution \nof incomes within loan application outcomes differs between genders \nand reflects systemic inequality in approvals or loans, incomes, or both.", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017")
```


In exploring whether loan approvals processes systemically disadvantage any groups, I have chosen to explore gender as well as race. This graph shows the differences in average income for males and females within each outcome group (approved, denied, or other). Men who are approved for loans typically have higher incomes than women who are approved for loans. Men who are denied also have higher incomes, on average than women who are denied a loan. The other category contains people who did not provide information needed to complete their loan applications, and also reflects the societal differences in incomes by gender. 
```{r, error = FALSE, message = FALSE, warning = FALSE, fig.align = "center", fig.height=10, fig.width=11}
#loan application outcome rates by race and year (facet_grid)

hmda_ny_5yrs <-  hmda_ny_5yrs %>% mutate(income_bins = ifelse(applicant_income_000s %in% c(1:73), "Income Quartile 1", 
                                             ifelse(applicant_income_000s %in% c(74:108), "Income Quartile 2", 
                                                    ifelse(applicant_income_000s %in% c(109:180), "Income Quartile 3", 
                                                           ifelse(applicant_income_000s %in% c(181:133549), "Income Quartile 4", NA)))))


#total loans per quartile, year, race
quartile_race_year <- hmda_ny_5yrs %>% 
  group_by(income_bins, race_alternative, as_of_year, county_name) %>% 
  summarise(total = n())

quartile_race_year$race_alternative[which(quartile_race_year$race_alternative == "Black or African American")] <- "Black"

quartile_race_year_outcome <- hmda_ny_5yrs %>% 
  group_by(income_bins, race_alternative, outcome, as_of_year, county_name) %>% 
  summarise(per_outcome = n())

quartile_race_year_outcome$race_alternative[which(quartile_race_year_outcome$race_alternative == "Black or African American")] <- "Black"

quartile_race_year_outcome %>% 
  left_join(quartile_race_year, 
            by = c("income_bins", "race_alternative", "as_of_year", "county_name")) %>% 
  mutate(outcome_rate = per_outcome/total) %>% 
  filter(race_alternative %in% c("Asian", "White", "Black", "Other"), county_name == "Kings County", !is.na(income_bins)) %>% 
  ggplot() + geom_line(aes(x = as_of_year, y = outcome_rate, color = outcome)) + 
  facet_wrap(race_alternative~income_bins) + 
  scale_color_manual(values = c('#1f8a70', '#E8291A', '#FFE11A')) + 
  theme(text = element_text(family = "Meiryo"), 
        panel.background = element_rect(fill = '#F2F5F2'), panel.spacing = unit(1.5, "lines"), legend.position = "bottom", axis.ticks = element_blank()) + 
  labs(title = "Loan application approval rates for low-to-middle income Blacks in Brooklyn has been decreasing, \nwhile most other races and quartiles have steady approval rates", subtitle = "Rates over time broken down by income quartiles and by race, show that for \nWhites and Asians, approvals have been steadily increasing, \nwhile for Blacks and Others in the first quartile, rates are decreasing ", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017", color = "Outcome", x = "Year", y = "Rate")
```


By breaking down applicants into income quartiles, and by race, we can look at differences in approvial rates overtime, and identify the groups that are struggling with the changing NYC housing landscape. Approval rates for Black applicants in the first income quartile have been steadily decreasing over the last five years. Many other income groups and racial groups have not seen any decreases in their approval rates over time. This graph also allows one to contrast how high White approval rates are compared to Black approval rates in the first income quartile. The Other category is a combination of American Indians, Native Alaskans, Native Hawaiians, and Pacific Islanders. The Other category has very different approval rates over time for each income quartile, and for the first income quartile, approval rates were below denial rates for most years. 







