---
title: "An Exploration of NYC Home Mortgage Disclosure Act Data"
output: 
  html_document: 
    code_folding: hide
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(maptools)
library(ggmap)
library(broom)
library(ggrepel)
library(statar)
library(sf)
library(treemapify)
library(ggridges)
```

```{r, error = FALSE, message = FALSE, warning = FALSE, include=FALSE, cache = TRUE}
#loading data and making preliminary changes 
options("scipen"=100, "digits"=4)

hmda13_ny <- read_csv("hmda13_ny.csv")
hmda14_ny <- read_csv("hmda14_ny.csv")
hmda15_ny <- read_csv("hmda15_ny.csv")
hmda16_ny <- read_csv("hmda16_ny.csv")
hmda17_ny <- read_csv("hmda17_ny.csv")

hmda_ny_5yrs <- rbind(hmda13_ny, hmda14_ny, hmda15_ny, hmda16_ny, hmda17_ny) %>%
  mutate(census_tract_number = gsub("\\.", "", census_tract_number)) %>% filter(county_name %in% c("Bronx County", "Kings County", "Queens County", "New York County", "Richmond County"))

hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(outcome = ifelse(action_taken_name %in% 
                            c("Application approved but not accepted",
                              "Loan originated", "Loan purchased by the institution"),"Approved", 
                          ifelse(action_taken_name %in%
                                   c("Application denied by financial institution",
                                     "Preapproval request denied by financial institution"),
                                 "Denied", "Other")))

hmda_ny_5yrs$applicant_race_name_1[which(hmda_ny_5yrs$applicant_race_name_1 %in%
                                           c("Information not provided by applicant in mail, Internet, or telephone application",
                                             "Not applicable"))] <- "Unknown"


hmda_ny_5yrs <- hmda_ny_5yrs %>% mutate(race_alternative = applicant_race_name_1)
hmda_ny_5yrs$race_alternative[which(hmda_ny_5yrs$race_alternative %in% 
                                      c("American Indian or Alaska Native", "Native Hawaiian or Other Pacific Islander"))] <- "Other"

hmda_ny_5yrs$race_alternative[which(hmda_ny_5yrs$race_alternative == "Black or African American")] <- "Black"

summary(hmda_ny_5yrs$applicant_income_000s)

hmda_ny_5yrs <-  hmda_ny_5yrs %>% mutate(income_bins = ifelse(applicant_income_000s %in% c(1:76), "Income Quartile 1", 
    ifelse(applicant_income_000s %in% c(77:116), "Income Quartile 2", 
        ifelse(applicant_income_000s %in% c(117:204), "Income Quartile 3", 
            ifelse(applicant_income_000s %in% c(205:133549), "Income Quartile 4",
                   "Unknown")))))

hmda_ny_5yrs <-  hmda_ny_5yrs %>% mutate(boro = 
        ifelse(county_name == "New York County", "Manhattan", 
            ifelse(county_name == "Kings County", "Brooklyn", 
              ifelse(county_name == "Queens County", "Queens", 
                     ifelse(county_name == "Bronx County", "Bronx",
                            ifelse(county_name == "Richmond County", 
                                   "Staten Island", NA))))))

hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(win_loan_amount = winsorize(loan_amount_000s, probs = c(.01, .99)))

hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(win_income = winsorize(applicant_income_000s, probs = c(.01, .99)))

hmda_ny_5yrs <- hmda_ny_5yrs %>% mutate(win_loan_income_ratio = win_loan_amount/win_income)

#regroup denial reasons 
hmda_ny_5yrs$denial_reason_alt <- hmda_ny_5yrs$denial_reason_name_1

hmda_ny_5yrs$denial_reason_alt[which(hmda_ny_5yrs$denial_reason_alt %in% 
          c("Employment history", "Insufficient cash (downpayment, closing costs)","Mortgage insurance denied","Unverifiable information"))] <- "Other"

census_tracts <- st_read("/Volumes/SPACESHIP/data_viz/data_scripts/visualizing-nyc-housing/2010_Census_Tracts", "geo_export_6e8e8432-67b3-4029-a002-fa59bbd3d5d3")

nta_shape <- st_read("/Volumes/SPACESHIP/data_viz/data_scripts/visualizing-nyc-housing/Neighborhood_Tabulation_Areas", "geo_export_f14927c2-d9cb-48d6-b521-916e745df67d")

rm(hmda13_ny, hmda14_ny, hmda15_ny, hmda16_ny, hmda17_ny)

crosswalk<- read_csv("crosswalk_nta_ct.csv")
crosswalk$X4 <- NULL

hmda_ny_5yrs <- hmda_ny_5yrs %>% left_join(crosswalk, by = c("census_tract_number" = "census_tract")) 

rm(crosswalk)
```


#Overview 

Homeownership is one of the most important ways of acculating wealth. In New York City, property values are high and thus homeownership is out of reach for many.Using Home Mortgage Disclosure Act data, I explore the demographics of mortgage applicants and the various outcomes of their applications in order to better understand who receives loans and who does not. Looking at how outcomes change over time and by group allows us to better understand the way opportunities within the city are changing.

##Loan Applicant Characteristics 

The following graphs breakdown applicants by race, income, and gender, to understand what type of people are applying for loans in New York City. 
```{r,error = FALSE, message = FALSE, warning = FALSE,  fig.width=10, fig.height=9, fig.align="center"}
#Loan applications are split pretty evenly between Manhattan, Brooklyn, and Queens, with White applicants representing a large share of the applications. The Unknown category is made up of applicants who did not disclose their race in their loan applications. Disparities in incomes between boroughs are shown in the second graph. 

#treemap for counties and race 
data_tree <- hmda_ny_5yrs %>% group_by(boro, race_alternative) %>% 
  summarise(n = n())

ggplot(data_tree, aes(area =n, label = race_alternative, subgroup = boro, fill = race_alternative)) +
  geom_treemap() + 
  geom_treemap_subgroup_border(color = '#F2F5F2') + 
  geom_treemap_text(place= "bottomright", color = '#F2F5F2',
                    fontface = "italic") + 
  geom_treemap_subgroup_text(place = "topleft", color = '#F2F5F2') + 
  scale_fill_manual(values = 
                      c('#FD7400', '#BF4182', '#580F45', '#0D177F', '#1f8a70')) +
  labs(title = "Loan Applicants Organized by Borough and \nRace HMDA Data 2013-2017") + 
  theme(text = element_text(family = "Meiryo"), 
        legend.position = "none", plot.title = element_text(size = 18))

data_tree <- hmda_ny_5yrs %>% filter(income_bins != "Unknown") %>% group_by(boro, income_bins) %>% 
  summarise(n = n())

ggplot(data_tree, aes(area =n, label = income_bins, subgroup = boro, fill = income_bins)) +
  geom_treemap() + 
  geom_treemap_subgroup_border(color = '#F2F5F2') + 
  geom_treemap_text(place= "bottomright", color = '#F2F5F2',
                    fontface = "italic") + 
  geom_treemap_subgroup_text(place = "topleft", color = '#F2F5F2') + 
  scale_fill_manual(values = 
                      c('#FD7400', '#BF4182', '#580F45', '#0D177F')) +
  labs(title = "Loan Applicants Organized by Borough and \nIncome Quartiles HMDA Data 2013-2017") + 
  theme(text = element_text(family = "Meiryo"), 
        legend.position = "none", plot.title = element_text(size = 18))
```


```{r, error = FALSE, message = FALSE, warning = FALSE, fig.align="center"}
#looking at differences in approval rates by male/female
hmda_ny_5yrs <- hmda_ny_5yrs %>% 
  mutate(win_income = winsorize(applicant_income_000s, probs = c(.01, .99)))

#filter by people without co-applicants
ggplot(filter(hmda_ny_5yrs, applicant_sex_name %in% c("Female","Male"), co_applicant_ethnicity_name == "No co-applicant"),
       aes(y=win_income, x=applicant_sex_name, fill=outcome), alpha = .3) +
  geom_boxplot(outlier.size = .5) + scale_y_log10(expand=c(0,0)) + 
  scale_fill_manual(values = c('#1f8a70', '#E8291A', '#FFE11A')) + 
  theme(text = element_text(family = "Meiryo"), 
        panel.background = element_rect(fill = '#F2F5F2'),
        plot.subtitle = element_text(face = "italic"),
        panel.grid.minor = element_blank(),
        panel.grid.major = element_line(color = "#404040"),
        plot.title = element_text(size = 18)) +
  labs(fill = "Outcome", x = "Gender of Applicant", 
       y = "Income (in $1,000s, Log Scale)", 
       title = "On Average, Male Applicants have Higher Incomes \nthan Female Applicants", 
       subtitle = "Income by Gender, grouped by Loan Outcome", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017") 
```


The graph above showcases gender disparities in loan applications. Men apply with higher incomes, and their average income regardless of the outcomes of their applications are higher than those of women applicants. 

```{r, error = FALSE, message = FALSE, warning = FALSE}
#ridges plot of loan to income ratio
breaks = c(seq(0, 15, by=2.5))

ggplot(hmda_ny_5yrs %>% filter(co_applicant_ethnicity_name== "No co-applicant"), aes(x = win_loan_income_ratio, y = race_alternative,
          fill = race_alternative)) +
  geom_density_ridges() + geom_vline(xintercept = 2.5) + scale_x_continuous(limits = c(0, 15), breaks = breaks) + 
  theme(plot.title = element_text(size = 18), 
        plot.subtitle = element_text(face = "italic"),
        text = element_text(family = "Meiryo"),
        panel.background = element_rect(fill = '#F2F5F2'),
        panel.grid.minor = element_blank(), 
        axis.ticks = element_blank()) + 
  scale_fill_manual(values = 
                      c('#FD7400', '#BF4182', '#580F45', '#0D177F', '#1f8a70')) + guides(fill = FALSE) + labs(title = "Historically Disadvantaged Groups (Black, Other) \nhave a Higher Concentration of \nLoan-to-Income Ratios over 2.5", subtitle = "Investopedia states that a loan amount below 2.5 times \nan applicant's annual income is affordable to the applicant", x = "Loan-to-Income Ratio", y = "Race", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017")
```


Despite not being able to calculate payment-to-income (PTI) ratio, a commonly used metric for understanding loan affordability, here I calculate a loan-to-income ratio. The distribution or ratios for disadvantaged groups (Black and Other) shows a greater density of ratios above 2.5, while White applicants have a more bell-curve shaped density. 

```{r, error = FALSE, message = FALSE, warning = FALSE, fig.height = 10}
#differences in outcome by loan applications, income, loan amount, race 

hmda_ny_5yrs %>% 
  filter(co_applicant_ethnicity_name== "No co-applicant", race_alternative != "Unknown", win_loan_amount <1000, win_income <200, outcome == "Approved") %>% 
  ggplot(aes(x = win_loan_amount, y = win_income)) + 
  geom_point(alpha = .2, color = 'black') + 
  geom_smooth(color = '#1f8a70') + 
  facet_wrap(~race_alternative) + 
  scale_x_continuous(breaks = c(0,250,350,500,750,1000)) + 
  geom_hline(yintercept = 100, color = '#FD7400') + 
  geom_vline(xintercept = 350, color = '#FD7400') + 
  labs(title = "Trends in Loan Amounts and Applicant Income \nsuggest that for Applicants with the Same Income, \nWhites Receive Smaller Loans than other Races", subtitle = "Approved Loans for Applicants with Incomes below $200,000 and \nLoan Amounts below $1,000,000, by Race \n\nFor comparison, the orange lines show an income of $100,000 and \na loan amount of $350,000. On Average, a White applicant \nwith an income of $100,000 is receiving a $350,000 loan, \nhowever, a non-white applicant with same income \nis receiving a much larger loan", 
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017",
       y = "Applicant Income", x = "Loan Amount") + 
  theme(legend.position = "bottom", plot.title = element_text(size = 18), 
        plot.subtitle = element_text(face = "italic"),
        text = element_text(family = "Meiryo"), 
        panel.grid.minor = element_blank(),
        panel.background = element_rect(fill = '#F2F5F2'),
        axis.ticks = element_blank())
```


The graph above tries to understand the relationship between loan amount and income through a localized regression line. It shows how for an applicant with a $100,000 income, the associated loan amount is different for each race. White applicants receive, on average, a \$250,000 loan, while all other groups receive higher loan values. This may suggest that White applicants are able to have higher downpayments, and thus lower loan amounts, or that historically disadvantaged groups seek less afforable loan amounts. 

##Outcomes of Loan Applications

```{r, error = FALSE, message = FALSE, warning = FALSE, fig.align="left", include = FALSE, eval = FALSE}
#approval rates by nta
#number of loans apps by nta
apps_by_nta <- hmda_ny_5yrs %>% 
  group_by(ntaname) %>% 
  summarise(total_apps = n())

#approvals by nta
approval_counts_by_nta <- hmda_ny_5yrs %>% 
  filter(outcome == "Approved") %>% 
  group_by(ntaname) %>% 
  summarise(total_approved = n())

approval_rates_by_nta <- 
  approval_counts_by_nta %>% 
  left_join(apps_by_nta, by = "ntaname") %>% 
  mutate(approval_rate = total_approved/total_apps*100)

nta_merged <- left_join(nta_shape, approval_rates_by_nta,
    by = c("ntaname")) %>% 
      st_as_sf()

#dealing with ggplot
nta_merged$breaks <- cut(nta_merged$approval_rate, c(40, 50, 60, 70, 80)) 

#plot- approval rates by census block #### 
ggplot() + 
  geom_sf(data = nta_merged,
               aes(fill = breaks), color = "white", lwd = 0) +
  labs(title = "Most Neighborhoods have Loan Approval \nRates between 50% and 60% ", subtitle= "Percent of Loan Applications Approved by Neighborhood Tabulation Area (NTA)",
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017") + 
  theme(axis.text=element_blank(),                      
        axis.title=element_blank(),                      
        panel.background = element_rect(fill = "#ffffff", color = NA),
        axis.line = element_blank(),
        plot.subtitle = element_text(face = "italic"),
        legend.position = c(0.25,.75),
        text = element_text(family = "Meiryo"), 
        plot.title = element_text(size = 18),
        axis.ticks = element_blank()) + guides(fill=guide_legend(title="Percent of \nApplications \nApproved (%)")) + scale_fill_manual(values = c("#fef0d9", "#fdcc8a", "#fc8d59", "#d7301f"))
```

An average, for the city as a whole, 65% of loan applications are approved. The following graphs looks at how this percent differs between racial groups and neighborhoods.  

```{r, error = FALSE, message = FALSE, warning = FALSE, fig.align = "center", fig.height= 15}
#approval rates by race by nta facet map with race
approval_counts_by_race <- hmda_ny_5yrs %>% 
  filter(outcome == "Approved") %>% 
  group_by(ntaname, applicant_race_name_1) %>% 
  summarise(total_approved = n())

df_for_chloropleth_race_approval <- hmda_ny_5yrs %>% 
  group_by(ntaname, applicant_race_name_1) %>% 
  summarise(total_apps = n()) %>% 
  left_join(approval_counts_by_race, 
            by = c("ntaname", "applicant_race_name_1")) %>%
  mutate(approval_rate = total_approved/total_apps*100)

nta_tract_merged <- left_join(df_for_chloropleth_race_approval,nta_shape,
    by = "ntaname") %>% 
      st_as_sf()

#dealing with ggplot
nta_tract_merged$breaks <- cut(nta_tract_merged$approval_rate, c(0, 20, 40, 60, 80, 100)) 

#plot- approval rates by census block #### 
ggplot(data = nta_tract_merged) + 
  geom_sf(data = nta_tract_merged, aes(fill = breaks), color = "white", lwd = 0) + facet_wrap(~applicant_race_name_1, ncol = 2) + 
  labs(title = "Asians, Whites, and Unknown Group have \nHigh Loan Approval Rates across New York City", subtitle = "Percent of Loan Applications Approved by Neighborhood Tabulation Area (NTA) by Race",
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017") + 
  theme(axis.text=element_blank(),                      
        axis.title=element_blank(),                      
        panel.background = element_rect(fill = "#ffffff", color = NA),
        axis.line = element_blank(),
        legend.position = "bottom",
        plot.subtitle = element_text(face = "italic"),
        text = element_text(family = "Meiryo"), 
        plot.title = element_text(size = 18),
        axis.ticks = element_blank()) + guides(fill=guide_legend(title="Percent of Applications Approved (%)")) + scale_fill_manual(values = c("#fef0d9", "#fdcc8a", "#fc8d59", "#e34a33", "#b30000"))

```


The graph above shows a clear difference in loan rates by race for different neighborhoods. Black applicants have approval rates between 40-60% in most neighborhoods, while White and Asian applicants have approval rates between 80-100% in most neighborhoods. 


```{r, error = FALSE, message = FALSE, warning = FALSE, fig.align = "center", fig.height=10, fig.width=11}
#loan application outcome rates by race and year (facet_grid)

#total loans per quartile, year, race
quartile_race_year <- hmda_ny_5yrs %>% filter(income_bins != "Unknown") %>% 
  group_by(income_bins, race_alternative, as_of_year, county_name) %>% 
  summarise(total = n())

quartile_race_year$race_alternative[which(quartile_race_year$race_alternative == "Black or African American")] <- "Black"

quartile_race_year_outcome <- hmda_ny_5yrs %>% 
  filter(income_bins != "Unknown") %>% 
  group_by(income_bins, race_alternative, outcome, as_of_year, county_name) %>% 
  summarise(per_outcome = n())

quartile_race_year_outcome$race_alternative[which(quartile_race_year_outcome$race_alternative == "Black or African American")] <- "Black"

highlights <- data.frame(income_bins =c("Income Quartile 1"), race_alternative =c("Black", "Other"))

quartile_race_year_outcome %>% 
  left_join(quartile_race_year, 
            by = c("income_bins", "race_alternative", "as_of_year", "county_name")) %>% 
  mutate(outcome_rate = per_outcome/total*100) %>% 
  filter(race_alternative %in% c("Asian", "White", "Black", "Other"),
         county_name == "Kings County", !is.na(income_bins)) %>% 
  ggplot() + geom_line(aes(x = as_of_year, y = outcome_rate, color = outcome)) + 
  facet_wrap(race_alternative~income_bins) + 
  scale_color_manual(values = c('#1f8a70', '#E8291A', '#FFE11A')) + 
  theme(text = element_text(family = "Meiryo"), 
        plot.subtitle = element_text(face = "italic"),
        plot.title = element_text(size = 18),
        panel.background = element_rect(fill = '#F2F5F2'), 
        panel.spacing = unit(1.5, "lines"), legend.position = "bottom", axis.ticks = element_blank()) + 
  labs(title = "Loan Approval Rates in Brooklyn for Low-to-Middle Income Blacks \nand Other Group has been Decreasing, while most other Races and Quartiles \nhave Steady Rates", 
       subtitle = "Approval Rates per Year by Income Quartiles and Race",
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017",
       color = "Outcome", x = "Year", y = "Percent of Applications (%)") + 
  geom_rect(data = highlights, 
            aes(xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=Inf), 
            fill = '#0D177F', alpha= .2)
```


```{r, error = FALSE, message = FALSE, warning = FALSE}
#approval rates for low income applicants, by NTA faceted by year 
apps_outcome_Q1_nta <- hmda_ny_5yrs %>% 
  filter(income_bins == "Income Quartile 1") %>% 
  group_by(ntacode, as_of_year, outcome) %>% 
  summarise(outcome_tot = n())

apps_total_Q1_nta <- hmda_ny_5yrs %>% 
  filter(income_bins == "Income Quartile 1") %>% 
  group_by(ntacode, as_of_year) %>% 
  summarise(tot_apps = n())

approval_rate_5yr <- apps_outcome_Q1_nta %>% 
  left_join(apps_total_Q1_nta, by = c("ntacode", "as_of_year")) %>%
  filter(outcome == "Approved") %>% 
  mutate(approval_rate = outcome_tot/tot_apps*100)

approval_rate_5yr$breaks <- cut(approval_rate_5yr$approval_rate, c(0, 20, 40, 60, 80, 100))

#need to merge back with census hmda and then merge that with census tracts
approval_rate_5yr %>% 
  left_join(nta_shape, by = "ntacode") %>% 
  ggplot() + geom_sf(aes(fill = breaks), lwd = 0) + 
  facet_wrap(~as_of_year) + scale_fill_manual(values = c('#ffffd4', '#fdd2c0', '#f3a683', '#e27a48', '#cc4c02')) + 
  guides(fill=guide_legend(title="Approval Rate (%)")) +
  theme(axis.text=element_blank(),                      
        axis.title=element_blank(),                      
        panel.background = element_rect(fill = "#ffffff", color = NA),
        axis.line = element_blank(),
        legend.position = c(.8, .3), 
        legend.background = element_blank(),
        plot.title = element_text(size = 18), 
        plot.subtitle = element_text(face = "italic"),
        text = element_text(family = "Meiryo"), 
        axis.ticks = element_blank()) + labs(title ="Approval Rates for Low-to-Middle Income Applicants \nDecline in most Boroughs, while Rates on Queens \nand Staten Island remain High", subtitle = "Approval Rates for Applicants with Incomes below $76,000/year", caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017")
```

Low-to-middle class residents are having a harder time remaining in the city as rents and property values rise. The graph above shows how approval rates between 40-60% for low-to-middle-income applicants were once spread across the city, and are now concentrated in a few neighborhoods and boroughs, including Queens and Staten Island. 


The following graph provides an understanding of why different groups are being denied for loans. Historically disadvantaged groups, such as Black and Other applicants, have a much higher chance of being denied because of their credit history, while White and Asian applicants are more frequently denied for an insufficient debt-to-income ratio. 

```{r, error = FALSE, message = FALSE, warning = FALSE, fig.height=6}

num_denied_by_race <- hmda_ny_5yrs %>% 
  filter(outcome == "Denied", !is.na(denial_reason_alt)) %>%
  group_by(race_alternative) %>% 
  summarise(tot_apps_denied = n())

denial_reasons_race <- hmda_ny_5yrs %>% 
  filter(outcome == "Denied", !is.na(denial_reason_alt)) %>% 
  group_by(race_alternative, denial_reason_alt) %>% 
  summarise(denied_by_reason = n())

#denial reasons rates race ####
denial_reasons_race %>% 
  left_join(num_denied_by_race, by = "race_alternative") %>%
  mutate(denial_rate = denied_by_reason/tot_apps_denied) %>% 
  ggplot() + 
  geom_bar(aes(x = race_alternative, 
               y = denial_rate, 
               fill = denial_reason_alt),
           stat = "identity") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        text = element_text(family = "Meiryo"), 
        panel.background = element_rect(fill = '#F2F5F2'), 
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        plot.title = element_text(size = 16),
        panel.grid = element_blank()) +
  scale_x_discrete(expand = c(0,0)) + scale_y_continuous(expand = c(0,0)) +
  labs(title = "Black Applicants and \nOther (Native American and Pacific Islanders) \nApplicants Denials are more frequently \ndue to Credit History, while this is an Infrequent Denial \nReason for Whites and Asians", 
       subtitle = "Distribution of Denial Reasons by Race", 
       caption = "Home Mortgage Disclosure Act Data, CFPB, 2013-2017", 
       x = "Applicant Race", y = "Denial Rate", fill = "Denial Reason") + scale_fill_discrete(name = "Denial Reason")  + scale_fill_manual(values = c('#FFE11A', '#004358', '#1f8a70','#FD7400',  '#0D177F'))
```